---
title: "GROW Molecular Formula Relationships"
author: "RED"
date: "2023-11-22"
output: html_document
---

As opposed to the other Rmarkdown which was looking at relationships from various cheminformatic derived information sets, this script is going directly correlate molecular formula intensities/relative intensities to the 7 carbon use classes. In order to accomplish this, we need to ensure that our intensities are normalized - as such, we will be using the pMartR package to identity the best normalization approach (Thompson et al., 2021 - Rapid Comm. Mass. Spec.).

Given the sensitivity of this approach, I wanted it to be broken out into it's own Rmarkdown rather than in the larger analytical one; this relies on the merged dataset generated by the other Rmarkdown, however.

https://analyticalsciencejournals.onlinelibrary.wiley.com/doi/epdf/10.1002/rcm.9068

## Setup
Loading packages and whatnot

```{r setup, include=FALSE}
# organizational packages
library(readxl)
library(tidyverse)
library(ggpubr)
library(ggthemes)
library(ggrepel)
library(gridExtra)
library(Rfast)

# analytical
library(Hmisc)
library(pmartR)

# functions
johnston_set = function(mol){
  # Johnston SE et al., 2021 - Biogeochemistry
  # https://link.springer.com/article/10.1007/s10533-021-00876-7#Sec2
  
  # empty object
  johnston_set = rep(NA, nrow(mol))
  
  # boundaries
  johnston_set[which(mol$AI_Mod < 0.5 & 
                       mol$HtoC_ratio < 1.5 & 
                       mol$OtoC_ratio < 0.5)] = "Highly Unsaturated and Phenolic, low O/C"
  johnston_set[which(mol$AI_Mod < 0.5 & 
                       mol$HtoC_ratio < 1.5 & 
                       mol$OtoC_ratio >= 0.5)] = "Highly Unsaturated and Phenolic, high O/C"
  johnston_set[which(mol$HtoC_ratio >= 1.5, mol$N == 0)] = "Aliphatic"
  johnston_set[which(mol$AI_Mod >= 0.67)] = "Condensed Aromatic"
  johnston_set[which(mol$AI_Mod > 0.5 & mol$AI_Mod < 0.67)] = "Polyphenolic"
  johnston_set[which(mol$HtoC_ratio >= 1.5, mol$N > 0)] = "Peptide-like"
  johnston_set[which(mol$OtoC_ratio > 0.9)] = "Sugar-like"
  
  # write out
  return(johnston_set)
  
} # could be case_when, but nervous

roebuck_set = function(mol){
  # Roebuck JA et al., 2022 - Geophyical Research Letters
  # https://agupubs.onlinelibrary.wiley.com/doi/full/10.1029/2022GL099535
  # modified from Seidel et al., 2015 - Marine Chemistry
  # https://www.sciencedirect.com/science/article/pii/S0304420315300049
  
  # empty object
  roebuck_set = rep(NA, nrow(mol))
  
  # boundaries
  roebuck_set[which(mol$AI_Mod >= 0.67 & mol$C >= 15)] = "Polycyclic Aromatic"
  roebuck_set[which(mol$AI_Mod >= 0.67 & mol$C >= 15 & 
                      (mol$N > 0 | mol$S > 0 | mol$P > 0))] = "Polycyclic Aromatic w/ Heteroatoms"
  roebuck_set[which(mol$AI_Mod >= 0.67 & mol$C < 15 &
                      (mol$N > 0 | mol$S > 0 | mol$P > 0))] = "High Aromaticity (<15C)"
  roebuck_set[which(mol$AI_Mod >= 0.5 & mol$AI_Mod < 0.67)] = "Highly Aromatic including Polyphenols"
  roebuck_set[which(mol$AI_Mod >= 0.5 & mol$AI_Mod < 0.67 &
                      (mol$N > 0 | mol$S > 0 | mol$P > 0))] = "Highly Aromatic w/ Heteroatoms"
  roebuck_set[which(mol$AI_Mod < 0.5 & mol$HtoC_ratio < 1.5)] = "Highly Unsaturated"
  roebuck_set[which(mol$AI_Mod < 0.5 & mol$HtoC_ratio < 1.5 &
                      (mol$N > 0 | mol$S > 0 | mol$P > 0))] = "Highly Unsaturated w/ Heteroatoms"
  roebuck_set[which(mol$HtoC_ratio >= 1.5 & mol$HtoC_ratio < 2)] = "Unsaturated Aliphatics"
  roebuck_set[which(mol$HtoC_ratio >= 1.5 & mol$HtoC_ratio < 2 &
                      (mol$N > 0 | mol$S > 0 | mol$P > 0))] = "Unsaturated Aliphatics w/ Heteroatoms (including peptides)"
  roebuck_set[which(mol$HtoC_ratio >= 2 & mol$OtoC_ratio < 0.9)] = "Saturated Compounds (including lipids)"
  roebuck_set[which(mol$HtoC_ratio >= 2 & mol$OtoC_ratio < 0.9 &
                      (mol$N > 0 | mol$S > 0 | mol$P > 0))] = "Saturated Compounds w/ Heteroatoms"
  roebuck_set[which(mol$HtoC_ratio >= 2 & mol$OtoC_ratio >= 0.9)] = "Saturated Compounds (including carbohydrates)"
  roebuck_set[which(mol$mol$HtoC_ratio >= 2 & mol$OtoC_ratio >= 0.9 &
                      (mol$N > 0 | mol$S > 0 | mol$P > 0))] = "Carbohydrate-like w/ Heteroatoms (including amino sugars)"
  
  # write out
  return(roebuck_set)
}
# import lambda functions
source("~/Documents/Code Information/FTICR Scripts/R Functions/getLambda.R")

# knit options
knitr::opts_chunk$set(echo = TRUE)

# define colors
color = c(`chemolithoautotroph` = "#8CADD2",
          `heterotroph-aromatic` = "#198035",
          `heterotroph-polymer` = "#093614",
          `heterotroph-sugar` = "#1F7C37",
          `methyl_c1` = "#7D52A7",
          `SCFA  and alcohol conversions` = "#7A82BA",
          `heterotroph-ch4` = "#72006D")
color = color[order(names(color))]

```

## Loading
Loading in the data

```{r load-in}
# set working directory (redunant)
setwd( "~/Documents/PNNL Analyses/GROW Paper/")

# load in merged data
data = read.csv("Merged_S19S_Wat_8.12_Data.csv")
mol = read.csv("Merged_S19S_Wat_8.12_Mol.csv")

# load in metadata
meta = read_xlsx("Table S1 - S19S Metadata.xlsx")

# load in carbon use
carbon.use = read_xlsx("carbon_usage.xlsx")

```

## SPANS and normalization
This section will use SPANS to identify the best normalization approach for this dataset and then apply the selected normalization.

```{r normalization}
# convert to pmartR
edata = data
emeta = mol
fdata = meta

# pmartR object
pmartR.obj = as.pepData(
  e_data = edata,
  f_data = fdata,
  e_meta = emeta,
  edata_cname = "X",
  fdata_cname = "ID",
  emeta_cname = "X",
  data_scale = "abundance",
  data_types = "Negative Ion"
)

# setting grouping (might be an issue...)
pmartR.obj = group_designation(pmartR.obj, main_effects = "ID")

# convert to log
attr(pmartR.obj, 'data_info')$data_scale = "log"

# run SPANS
spans.res = spans_procedure(pmartR.obj)

# clean up
rm(edata, emeta, fdata)

### SPANS revealed nothing useful due to experimental design.
### Proceeding with TSS.

# selecting assigned formulas only
data = data[!is.na(mol$MolForm),]
mol = mol[!is.na(mol$MolForm),]

# relative intensity
data = as.data.frame(
  apply(data.frame(data, row.names = 1), 2, function(x) (x/sum(x))*100)
)



```

Well, I was unaware of this requirement for SPANS - it requires an a priori defined grouping variable which we do not have in this situation. We might be able to revisit this if we want to normalize within something defined a posteriori (like stream order), so I'm leaving the code here. With this in mind, I'll be normalizing using TSS (e.g., relative intensity) for the assigned formulas only.

I might revisit the normalization with the "pre-merged" data at a later date.


## Molecular Formula-to-Guild Relationships
This section is running the Spearman correlations comparing each guild to each molecular formula. I'm a bit worried that the sheer number of pairwise comparisons is going to destroy any statistical power that we have, but we will see...

```{r correlations}
# select only assigned formulas
data = data[!is.na(mol$MolForm),]
mol = mol[!is.na(mol$MolForm),]

# removing duplicate molecular formulas
# some peaks can be assigned the same formula
data = data %>% mutate(MolForm = mol$MolForm) %>%
  group_by(MolForm) %>%
  summarise_all(.funs = "sum")
data = data.frame(data, row.names = 1)
mol = mol[!duplicated(mol$MolForm),]

# ensuring identical order
data = data[order(row.names(data)),]
mol = mol[order(mol$MolForm),]

# confirm matching data and mol
if(!identical(row.names(data), mol$MolForm)){
  stop("The FTICR-MS data objects do not having matching masses.")
}

# preparing objects for correlation
corr.data = t(data)
corr.guild = carbon.use %>% 
  pivot_wider(names_from = Guild, values_from = abund)
corr.guild[is.na(corr.guild)] = 0
corr.guild = as.matrix(data.frame(corr.guild, row.names = 1, check.names = F))

# ensuring matching objects
corr.data = corr.data[which(row.names(corr.data) %in% row.names(corr.guild)),]
corr.data = corr.data[order(row.names(corr.data)),]
corr.guild = corr.guild[order(row.names(corr.guild)),]

# running correlations
molform_stats = rcorr(corr.guild, corr.data, type = "spearman")

# removing molform self-correlations
molform_stats$r = molform_stats$r[,-which(colnames(molform_stats$r) %in% colnames(corr.data))]
molform_stats$r = molform_stats$r[-which(row.names(molform_stats$r) %in% colnames(corr.guild)),]

molform_stats$P = molform_stats$P[,-which(colnames(molform_stats$P) %in% colnames(corr.data))]
molform_stats$P = molform_stats$P[-which(row.names(molform_stats$P) %in% colnames(corr.guild)),]

# creating finalized correlation object
rho = data.frame(molform_stats$r, check.names = F) %>%
  rownames_to_column("MolForm") %>%
  gather(Guild, rho, -MolForm)
pval = data.frame(molform_stats$P, check.names = F) %>%
  rownames_to_column("MolForm") %>%
  gather(Guild, pvalue, -MolForm)
final_stats = rho %>% left_join(pval, by = c("MolForm", "Guild"))

# remove failed correlations (likely both abundances = 0 or invariant)
final_stats = final_stats[!is.na(final_stats$pvalue),]

# p-value adjustment
final_stats$padj = p.adjust(final_stats$pvalue, method = "fdr")

# clean-up
rm(corr.data, corr.guild, molform_stats)

# write stats
write.csv(final_stats, "Correlations/Guild-MolForm_Relationships.csv", 
          quote = F)

```

# Looking at the results, briefly...
This section is just giving us a glimpse at the significant relationships - nothing too deep.

```{r data-glimpse}
# Counting significant relationships by guild
final_stats %>% filter(padj < 0.05) %>%
  mutate(Direction = case_when(rho > 0 ~ "Positive",
                               rho < 0 ~ "Negative")) %>%
  group_by(Guild, Direction) %>%
  summarise(Count = n()) %>%
  ggplot(aes(x = Guild, y = Count))+
  geom_bar(stat = "identity", aes(fill = Guild))+
  scale_fill_manual(values = color)+
  facet_grid(.~Direction)+
  theme_bw()+
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

### generating mol object describing each guild relationships
# calculating new boundary sets
mol$bs_johnston = johnston_set(mol)
mol$bs_roebuck = roebuck_set(mol)

# calculate lambda
get_comp = get_compositions(mol) # lambda requires specific elcomp formulation
lambda = as.data.frame(get_lambda(get_comp$chemical_compositions)) # calc lambda

# name lambda object
names <- rep("", 62)
names[1:12] <- c("delGcox0","delGd0","delGcat0","delGan0","delGdis0","lambda0",
                 "delGcox","delGd","delGcat","delGan","delGdis","lambda")
stoich_colnames <- c("donor","h2o","hco3","nh4","hpo4","hs","h","e","acceptor","biom")
stoich_types <- c("stoichD","stoichA","stoichCat","stoichAn","stoichMet")

for (i in 1:length(stoich_types)) {
  names[((i-1)*10+13):(i*10+12)] <- array(sapply(stoich_types[i], paste, stoich_colnames, sep="_"))
}

colnames(lambda) <- names

# rearrange lambda
lambda['MolForm'] <- get_comp$formulas
lambda = as.data.frame(lambda[,c("MolForm", "delGcox0", "delGcox", "lambda0", "lambda", "delGd0", "delGd")])
colnames(lambda) = c("MolForm","delGcox0PerCmol","delGcoxPerCmol", "lamO20","lamO2","delGd0","delGd")
lambda = lambda[!duplicated(lambda$MolForm),]

# add lamba to mol
mol = mol %>% left_join(lambda, by = "MolForm")

# cleanup
rm(names, stoich_colnames, stoich_types, get_comp, lambda)

# empty object
guild_mol = NULL

# looping through guilds to select molforms
for(curr.guild in unique(final_stats$Guild)){
  # identify significant
  temp = final_stats[final_stats$padj < 0.05,]
  
  # identify molforms
  temp.pos = temp$MolForm[which(temp$Guild %in% curr.guild &
                                  temp$rho > 0)]
  temp.neg = temp$MolForm[which(temp$Guild %in% curr.guild &
                                  temp$rho < 0)]
  
  # select molforms from mol
  if(length(temp.pos) == 1|length(temp.neg) == 1){
    temp = data.frame(mol[which(mol$MolForm %in% c(temp.pos, temp.neg)),],
                      Direction = case_when(length(temp.pos) == 1 ~ "Positive", 
                                            length(temp.neg) == 1 ~ "Negative"),
                      check.names = F)
  } else {
    temp = data.frame(mol[which(mol$MolForm %in% temp.pos),], 
                      Direction = "Positive", check.names = F)
    temp = rbind(temp,
                 data.frame(mol[which(mol$MolForm %in% temp.neg),], 
                            Direction = "Negative", check.names = F))
  }
  
  # associate formulas with guild
  temp$Guild = curr.guild
  
  # add into final object
  guild_mol = rbind(guild_mol, temp)
  
  # clean up
  rm(temp, temp.pos, temp.neg)
}

# plotting metrics
plot1 = guild_mol %>% 
  select(AI_Mod, DBE, NOSC, lamO2, Guild, Direction) %>%
  gather(Variable, Value, -Guild, -Direction) %>%
  ggplot(aes(x = Guild, y = Value))+
  geom_boxplot(aes(fill = Direction))+
  scale_fill_manual(values = c("white", "gray"))+
  facet_wrap(.~Variable, scales = "free_y")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
plot1

# plotting bailey boundary set
plot2 = guild_mol %>%
  group_by(Guild, bs2_class, Direction) %>%
  summarise(Count = n()) %>%
  ungroup() %>% group_by(Guild, Direction) %>%
  mutate(Rel_Count = (Count/sum(Count))*100) %>%
  gather(Variable, Value, -Guild, -bs2_class, -Direction) %>%
  ggplot(aes(x = Guild, y = Value))+
  geom_bar(stat = "identity", aes(fill = bs2_class))+
  scale_fill_gdocs(name = "Bailey Set")+
  facet_wrap(Direction~Variable, scales = "free_y")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
plot2

# plotting johnston boundary set
plot3 = guild_mol %>%
  group_by(Guild, bs_johnston, Direction) %>%
  summarise(Count = n()) %>%
  ungroup() %>% group_by(Guild, Direction) %>%
  mutate(Rel_Count = (Count/sum(Count))*100) %>%
  gather(Variable, Value, -Guild, -bs_johnston, -Direction) %>%
  mutate(bs_johnston = case_when(is.na(bs_johnston) ~ "Other",
                                 .default = bs_johnston)) %>%
  ggplot(aes(x = Guild, y = Value))+
  geom_bar(stat = "identity", aes(fill = bs_johnston))+
  scale_fill_solarized(name = "Johnston Set")+
  facet_wrap(Direction~Variable, scales = "free_y")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
plot3

# plotting roebuck boundary set
plot4 = guild_mol %>%
  group_by(Guild, bs_roebuck, Direction) %>%
  summarise(Count = n()) %>%
  ungroup() %>% group_by(Guild, Direction) %>%
  mutate(Rel_Count = (Count/sum(Count))*100) %>%
  gather(Variable, Value, -Guild, -bs_roebuck, -Direction) %>%
  mutate(bs_roebuck = case_when(is.na(bs_roebuck) ~ "Other",
                                 .default = bs_roebuck)) %>%
  ggplot(aes(x = Guild, y = Value))+
  geom_bar(stat = "identity", aes(fill = bs_roebuck))+
  scale_fill_stata(name = "Roebuck Set")+
  facet_wrap(Direction~Variable, scales = "free_y")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
plot4

# writing plots
ggsave("Figures/MolForm-Relationships_Metrics.pdf", plot1, 
       height = 12, width = 12)
ggsave("Figures/MolForm-Relationships_Bailey.pdf", plot2, 
       height = 12, width = 12)
ggsave("Figures/MolForm-Relationships_Johnston.pdf", plot3, 
       height = 12, width = 12)
ggsave("Figures/MolForm-Relationships_Roebuck.pdf", plot4, 
       height = 12, width = 12)


```